Biogeography and Environmental Conditions Shape Phage and Bacteria Interaction Networks Across the Healthy Human Microbiome

### 2017-03-31 doc/manuscript.Rmd
Changed repo name.
Updated with Jeniors edits.

Previous work has relied on studying bacterial and phage communities in isolation, reducing them to two separate communities. In reality, bacteria and phage communities are dynamic and complex, acting in concert to share genetic information and to maintain stable ecosystems, and removal of members can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. The conventional approach to understanding the human microbiome is unable to capture how these communities interact and influence each other, but relationship-based network approaches allow us to capture this information. We aimed to bridge this knowledge gap by characterizing human bacterial and phage communities by their relationships, leveraging machine learning and graph theory techniques. We focused on characterizing the network diversity of the human microbiome (including skin and gut) to provide a foundation for further studies of disease network dynamics.

We began characterizing human microbiome phage and bacteria networks by leveraged three published microbiome datasets with paired virus and bacterial metagenomic sequence sets [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. These datasets included gut and skin samples, allowing us to gain broader insights into human microbiome network diversity across different body sites. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Moebus:1981kp; @Flores:2013hc]. Our metagenomic interaction inference model was powered beyond previous models by its inclusion of protein interaction data, inclusion of negative interactions as well as positive, and the use of a more sophisticated machine learning algorithm instead of linear models [@Edwards:2015iz].

Using this approach, we were able to go beyond contemporary isolated methods to provide a basic understanding of the community network dynamics associated with healthy human phages and bacteria. By building and utilizing a microbiome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support communities with distinct communication structures and propensities toward community instability. By understanding the healthy state of network structures across the human body, we will empower future studies to begin investigating how these community structures change in disease or otherwise altered states.

@@

## Cohort Curation and Sample Processing
We studied the differences in microbiome phage-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, including a study of the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. The viromes associated with these datasets were all subjected to virus-like particle (VLP) purification to eliminate other organismal DNA including bacteria, fungi, and humans, thus allowing the original authors and ourselves to assess the actively replicating virome.

The publications we utilized were published over the span of five years using different methods and technologies, and therefore yielded different sequence abundances **(Supplemental Figure \ref{SequenceStats} A-B)**. The bacterial and viral sequences from those publications were quality filtered and assembled into contigs. Because contig assembly most commonly returns genome fragments, we clustered related bacteria and phage contigs by k-mer frequency and co-abundance using the CONCOCT algorithm. These clusters represent operationally defined units of related bacteria and phage genomes that we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht]. The resulting contigs and OGUs demonstrated high sequence coverage and length **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**.

@@

Four phage and bacterial genomic markers were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) CRISPR targeting of phages by bacterial CRISPR spacer sequences, and 4) similarity of protein families known to be associated with known protein-protein interactions [@Orchard:2014hq]. The resulting random forest model performed well, with an AUC of `r signif(genmodelper$ROC, digits = sigfig)`, a sensitivity of `r signif(genmodelper$Sens, digits = sigfig)`, and a specificity of `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was nucleotide similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, while CRISPRs were minimally important, likely because they were redundant with the BLAST information. Approximately one third of the relationships yielded no score and were therefore unknown, unable to be assigned an interaction **(Figure \ref{RocCurve} C)**.

We used our random forest model to classify the relationships between bacteria and phages in the experimental datasets. The relationships within the three studies were used to construct one master network, containing the three study sub-networks, which themselves each contained sub-networks for each sample **(Figure \ref{RocCurve} D)**. Metadata including study, sample ID, disease, and abundance within the community (based on sequence count) were also stored in the multi-study master network to allow for effective parsing and downstream analysis **(Supplemental Figure \ref{NetworkDiagram})**. The resulting master network was highly connected and contained `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], big.mark = ",")` infectious relationships among `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], big.mark = ",")` nodes, `r prettyNum(nestats[c(nestats$cats %in% "PhageNodes"),"values"], big.mark = ",")` of which represented phages and `r prettyNum(nestats[c(nestats$cats %in% "BacteriaNodes"),"values"], big.mark = ",")` that represented bacteria. Although the network was highly connected, not all relationships were present in all samples. Furthermore, relationships were weighted by relative abundance of their associated bacteria and phages, meaning lowly abundant relationships could be present but insignificant compared to those that were highly abundant. Like the master network, the skin network exhibited a diameter of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Diameter"], big.mark = ",")` (measure of graph size; the greatest number of traversed vertices required between two vertices) and included almost all of the master network nodes and edges **(Figure \ref{RocCurve} E - F)**. The gut diet and twin sample sets each contained fewer than 150 vertices, fewer than 20,000 relationships, and diameters of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "DietStudy"),"Diameter"], big.mark = ",")`, suggesting more sparsely related phages and bacteria **(Figure \ref{RocCurve} E - F)**.

## Individuality of Microbial Networks
Skin and gut community membership and diversity are highly personal, with people remaining more similar to themselves over time compared to other people [@Grice:2009ee; @Hannigan:2015fz; @Minot:2013ih]. We determined whether this personal conservation extended to network structure using our microbiome network model. We calculated the degree of dissimilarity between each subject's network based on phage and bacteria abundance, as well as centrality. Centrality was evaluated by first calculating the weighted eigenvector centrality of all bacteria and phages within each sample graph. Conceptually, this metric defines central phages as those that are highly abundant and infect many bacteria which themselves are abundant and infected by many other phages. Bacterial centrality was defined using the same metric. We then calculated the similarity of community networks using the weighted eigenvector centrality of all nodes between all samples. Biologically, samples with similar network structures are interpreted as having similar capacities to influence other microbes within the community, transmit genetic material (e.g. horizontal gene transfer), and maintain stability.

@@

We evaluated the diet-associated differences in gut network structure by quantifying how central each sample's network was on average. We accomplished this by utilizing two common centrality metrics: degree centrality and closeness centrality. Degree centrality, the simplest centrality metric, was defined as the number of connections each phage made to bacteria, or each bacterium made to phages. Because this metric alone offers only minimal insight, we supplemented it with measurements of closeness centrality. Closeness centrality is a metric of how close each phage or bacterium is to all of the other phages and bacteria in the network. A higher closeness centrality suggests that genetic information or the effects of altered abundance would be more impactful to all other microbes in the system. A network with higher average closeness centrality also indicates an overall greater degree of connections, which suggests a greater resilience to instability. Because these values are assigned to each phage and bacterium within each network, we calculated the average connectedness and corrected for the maximum potential degree of connectedness to obtain a single value for the connectedness of each graph.

@@

We demonstrated the diversity of relationships across the human body by showing that microbiome relationship structure, and therefore the potential for genetic transfer and stability maintenance, differs significantly between body sites and environmental conditions. We found that other environmental factors, such as diet, also influence relationship structures. This information builds on previous work by suggesting that bacteria and phages not only preferentially colonize different body sites, but also interact differently, allowing for different communication structures and capacities for maintaining stability. Our results will also provide a foundation for future studies to begin evaluating how microbiome network dynamics change in disease states, and how the information can be leveraged therapeutically.

@@

https://github.com/SchlossLab/Hannigan-2016-ConjunctisViribus.

>>>>

Previous work has relied on studying bacterial and phage communities in isolation, reducing them to two separate communities. In reality, bacteria and phage communities are dynamic and complex. They frequently share genetic information and work together to maintain stable ecosystems. Removal of members can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. The conventional approach to understanding the human microbiome has been unable to capture how these communities interact and influence each other, but relationship-based network approaches would allow us to capture this information. We aimed to bridge this knowledge gap by characterizing human bacterial and phage communities by their relationships, leveraging machine learning and graph theory techniques. We focused on characterizing the network diversity of the human skin and gut microbiomes to provide a foundation for further studies of disease network dynamics and to gain broader insights into human microbiome network diversity across different body sites.

We first investigated human microbiome phage and bacteria networks by utilizing three published microbiome datasets with paired virus and bacterial metagenomic sequence sets [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Moebus:1981kp; @Flores:2013hc]. This metagenomic interaction inference model was powered beyond previous models by its inclusion of protein interaction data, negative as well as positive interactions, and the use of a more sophisticated machine learning algorithm instead of simple linear models [@Edwards:2015iz].

Through this approach, we were able to go beyond contemporary isolated methods to provide a basic understanding of the community network dynamics associated with healthy human phages and bacteria. By building and utilizing a microbiome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support communities with distinct communication structures and propensities toward community instability. By understanding the healthy state of network structures across the human body, we will empower future studies to begin investigating how these community structures change in disease and their overall impact on human health.

@@

## Cohort Curation and Sample Processing
We studied the differences in microbiome phage-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, including a study of the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. The viromes associated with these datasets were all subjected to virus-like particle (VLP) purification to eliminate other organismal DNA and allow for direct assessment of the actively replicating virome.

The publications we utilized were published over the span of five years using different methods and technologies, and therefore yielded different sequence abundances **(Supplemental Figure \ref{SequenceStats} A-B)**. The bacterial and viral sequences from those publications were quality filtered and assembled into contigs. Because contig assembly most commonly returns genome fragments, we clustered related bacteria and phage contigs by k-mer frequency and co-abundance using the CONCOCT algorithm. We interpreted these clusters as operational units of related bacteria and phage genomes that we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht]. The resulting contigs and OGUs demonstrated high sequence coverage and length **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**.

@@

Four phage and bacterial genomic markers were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) CRISPR targeting of phages by bacterial CRISPR spacer sequences, and 4) similarity of protein families known to be associated with known protein-protein interactions [@Orchard:2014hq]. The resulting random forest model performed well, with an AUC of `r signif(genmodelper$ROC, digits = sigfig)`, a sensitivity of `r signif(genmodelper$Sens, digits = sigfig)`, and a specificity of `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was nucleotide similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, while CRISPRs were minimally important. Approximately one third of the relationships yielded no score and were therefore unknown, unable to be assigned an interaction **(Figure \ref{RocCurve} C)**.

Next we used our random forest model to classify the relationships between bacteria and phages in the experimental datasets. The relationships within the three studies were used to construct one master network, containing the three study sub-networks, which themselves each contained sub-networks for each sample **(Figure \ref{RocCurve} D)**. Metadata including study, sample ID, disease, and abundance within the community (based on sequence count) were also stored in the multi-study master network to allow for effective parsing and downstream analysis **(Supplemental Figure \ref{NetworkDiagram})**. The resulting master network was highly connected and contained `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], big.mark = ",")` infectious relationships among `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], big.mark = ",")` nodes, `r prettyNum(nestats[c(nestats$cats %in% "PhageNodes"),"values"], big.mark = ",")` of which represented phages and `r prettyNum(nestats[c(nestats$cats %in% "BacteriaNodes"),"values"], big.mark = ",")` that represented bacteria. Although the network was highly connected, not all relationships were present in all samples. Furthermore, relationships were weighted by relative abundance of their associated bacteria and phages, meaning lowly abundant relationships could be present but insignificant compared to those that were highly abundant. Like the master network, the skin network exhibited a diameter of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Diameter"], big.mark = ",")` (measure of graph size; the greatest number of traversed vertices required between two vertices) and included almost all of the master network nodes and edges **(Figure \ref{RocCurve} E - F)**. The gut diet and twin sample sets each contained fewer than 150 vertices, fewer than 20,000 relationships, and diameters of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "DietStudy"),"Diameter"], big.mark = ",")`, which indicated more sparsely related phages and bacteria **(Figure \ref{RocCurve} E - F)**.

## Individuality of Microbial Networks
Skin and gut community membership and diversity are highly personal, with people remaining more similar to themselves over time compared to other people [@Grice:2009ee; @Hannigan:2015fz; @Minot:2013ih]. We determined whether this personal conservation extended to network structure using our microbiome network model. To accomplish this, we calculated the degree of dissimilarity between each subject's network based on phage and bacteria abundance, as well as centrality. Weighted eigenvector centrality was calculated for all bacteria and phages within each sample graph. This metric defines central phages as those that are highly abundant and infect many distinct bacteria which themselves are abundant and infected by many other phages. Similarly, bacterial centrality was defined as those bacteria that were both abundant and susceptible to numerous phages which infected many bacteria. We then calculated the similarity of community networks using the weighted eigenvector centrality of all nodes between all samples. Samples with similar network structures were interpreted as having similar capacities for maintaining stability and transmitting genetic material.

@@

We evaluated the diet-associated differences in gut network structure by quantifying how central each sample's network was on average. We accomplished this by utilizing two common centrality metrics: degree centrality and closeness centrality. Degree centrality, the simplest centrality metric, was defined as the number of connections each phage made to bacteria, or each bacterium made to phages. We more completely quantified the network properties by supplementing degree centrality with measurements of closeness centrality. Closeness centrality is a metric of how close each phage or bacterium is to all of the other phages and bacteria in the network. A higher closeness centrality suggests that genetic information or the effects of altered abundance would be more impactful to all other microbes in the system. A network with higher average closeness centrality also indicates an overall greater degree of connections, which suggests a greater resilience to instability. Because these values are assigned to each phage and bacterium within each network, we calculated the average connectedness and corrected for the maximum potential degree of connectedness to obtain a single value for the connectedness of each graph.

@@

We demonstrated the diversity of relationships across the human body by showing that microbiome relationship structure, and therefore the potential for genetic transfer and stability maintenance, differs significantly between body sites and environmental conditions. We found that other environmental factors, such as diet, also influence relationship structures. This information builds on previous work by suggesting that bacteria and phages not only preferentially colonize different body sites, but also interact differently, allowing for different communication structures and capacities for maintaining stability. Our results will also provide a foundation for future studies to begin evaluating how microbiome network dynamics change in disease states, and how the information can be applied therapeutically.

@@

https://github.com/SchlossLab/Hannigan_ConjunctisViribus_GenRes_2017
